---
title: "Calling SCWorkflow from the CLI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calling SCWorkflow from the CLI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
```

SCWorkflow includes an executable file called `scworkflow`.
Any user-facing function in the SCWorkflow R package can be called with
`scworkflow [function]` from the unix CLI.
Function arguments are passed in via a JSON file.
In addition to arguments used by the function,
the JSON file can contain the following keys:

  - `object_input_rds` - file path to an existing object (Seurat object or SCWorkflow list) in RDS format. This is required if the SCWorkflow function has `object` as an argument (most user-facing functions do).
  - `object_output_rds` - file path to write the result to.

Run `scworkflow --help` to see the full CLI usage:

```{r help, echo=FALSE, results='asis'}
cat("```")
SCWorkflow:::cli_usage(con = stdout())
cat("```")
```

## Installing the SCWorkflow CLI


### Installation on a personal computer

After installing the R package, you can use `system.file()` to locate the
`scworkflow` executable file with R:

```{r install}
# remotes::install_github("CCBR/MOSuite", dependencies = TRUE)
system.file("exec", "scworkflow", package = "SCWorkflow")
```

You should add this executable to your `PATH` environment variable.

```sh
export PATH="$PATH:/path/to/exec/scworkflow"
```

## Example end-to-end script

You can create a shell script to run the full SCWorkflow pipeline.
This script assumes you have a directory `json_args/` with JSON files to set
each function's arguments.

```{r script_e2e, echo=FALSE, results='asis'}
cat("```bash\n")
cat(readr::read_lines(system.file("extdata", "example_script.sh", package = "SCWorkflow")), sep = "\n")
cat("\n```")
```

The example script and accompanying JSON files are included in the package data.
You can copy them to your working directory with R:

```{r example_data_paths, eval=FALSE}
# copy the example script
file.copy(system.file("extdata", "TestRunjson.sh", package = "SCWorkflow"), to = "./")
# copy the JSON files
file.copy(system.file("extdata", "json_args", package = "SCWorkflow"), to = "./", recursive = TRUE)
# copy the test data files
file.copy(system.file("extdata", "chariou", package = "SCWorkflow"), to = "./", recursive = TRUE)

singularity shell --bind $(pwd):/mnt docker://nciccbr/scworkflow:v1.0.2-galaxy
cd /mnt
R -e 'file.copy(system.file("extdata", "TestRunjson.sh", package = "SCWorkflow"), to = "./")'
R -e 'file.copy(system.file("extdata", "json_args", package = "SCWorkflow"), to = "./", recursive = TRUE)'
R -e 'file.copy(system.file("extdata", "chariou", package = "SCWorkflow"), to = "./", recursive = TRUE)'
```
Then run the script from the CLI:

```bash
bash ./TestRunjson.sh
```

The .rds output for each function can be set in the corresponding .json file

