---
title: "Differential Expression Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SCWorkflow-DEG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE, message = FALSE
)

library(Seurat)
library(data.table)
library(dplyr)
library(ggplot2)

run_Chunks=F
```

```{r,include=F,echo=F,eval=run_Chunks}

Anno_SO=readRDS("./images/CN_SO.rds")

```

## DE with Find Markers 

This function performs a DE (differential expression) analysis on a merged Seurat object to identify expression markers between different groups of cells (contrasts). This analysis uses the FindMarkers() function of the Seurat Workflow. 

**Methodology**
Differential expression analysis (DEG) is a fundamental technique in single-cell genomics research. The goal of DEG analysis is to identify genes that exhibit significant changes in expression levels between different groups of cells or conditions, thereby uncovering potential markers that distinguish these groups.
This function takes a merged Seurat object [1] as input, expected to contain single-cell data from multiple samples, along with relevant metadata and SingleR annotations, which provide information about cell identity. To perform the DEG analysis, the user can choose from various statistical algorithms, such as MAST [2], wilcox [3], bimod [4], and more, which accommodate different types of experimental designs and assumptions about the data. The user can control the sensitivity of the analysis by setting the minimum fold-change in gene expression between the groups to be considered significant. Additionally, users can specify the assay to be used for the analysis, whether it is the scaled data (SCT) or raw RNA counts.
For best results, it is recommended to use this function with well-curated and preprocessed single-cell data, ensuring that the Seurat object contains relevant metadata and SingleR annotations. Users should carefully select the samples and contrasts based on their experimental design and research questions. Additionally, exploring different statistical algorithms and adjusting the threshold can fine-tune the DEG analysis and reveal more accurate gene expression markers.

```{r,eval=run_Chunks}


          DEG_table=degGeneExpressionMarkers(object = Anno_SO,
                                   samples = c("CD8dep","Combo","ENT","NHSIL12","PBS"),
                                   contrasts = c("0-1"),
                                   parameter.to.test = "SCT_snn_res_0_2",
                                   test.to.use = "MAST",
                                   log.fc.threshold = 0.25,
                                   assay.to.use = "SCT",
                                   use.spark = F)

```
```{r}
if (eval_chunk==T) {
  write.table(DEG_table$df  ,
            file = "./images/DEG_Table.txt",
            sep = '\t',
            row.names = F)
}else{
  DEG_table=read.delim("./images/DEG_Table.txt")
}

  DEG_table%>%knitr::kable()

```

1. https://satijalab.org/seurat/
2. https://rglab.github.io/MAST/
3. Dalgaard, Peter (2008). Introductory Statistics with R. Springer Science & Business Media. pp. 99–100
4. https://en.wikipedia.org/wiki/Multimodal_distribution

## Pseudo Bulk Method

### Aggregate Seurat Counts

This function is the first step in a Pseudobulk analysis of your scRNA-seq dataset. It groups cells based on chosen categorical variable(s) in the Seurat Object's Metadata and aggregates the counts of each gene in each group.

The output is a table of aggregate expression in which the rows are genes and the columns are values found in the chosen Pseudobulk variable. If you select multiple categories to aggregate by (e.g. Category1: A,B,C and Category2: D,E,F), cells will be grouped by combinations of category variables (e.g. A_D, A_E, A_F, B_D, B_E, B_F). By default, gene counts are averaged across cells in each group.

```{r,eval=F}
aggregateCounts(object=so,
               var.group=var_group,
               slot=slot)
```


### Statistical Analysis using Limma

Given a matrix (typically log-normalized gene expression) and a metadata table, this will run one- and two-factor statistical analyses on groups using linear or mixed effects models with limma. Reference.

There are 2 ways for treating Donor or Patient - one as a random effect and the other as a fixed effect

When Using a Mixed Effects Model (Donor as random effect):

Add Donor column to the Donor Variable Column
Do not add the Donor variable under Covariate Columns. It will be handled separately in the Donor Variable Column as a random effect.
The Covariate Columns field should include any other variables except the Donor.
When Using a Basic Linear Model (Donor as fixed effect):

You can add the Donor column as a covariate under Covariate Columns, where it will be treated as a fixed effect.
Additional variables can be included under Covariate Columns
Ensure the Donor Variable Column is left blank.
This function is a Beta version and is undergoing active development. If you encounter problems, please contact CCBR NCICCBRNIDAP@mail.nih.gov


```{r,eval=F}
Pseudobulk_LimmaStats()
```

## Visualizations

### Volcano Plot - Enhanced

This function utilizes the *EnhancedVolcano* R Bioconductor package to generate publication-ready volcano plots for differential expression analyses, offering a number of customizable visualization options and optimizing gene label placement to avoid clutter


**Methodology**
A volcano plot is a type of scatterplot that shows statistical significance (P value) versus magnitude of change (fold change). It enables quick visual identification of genes with large fold changes that are also statistically significant. These may be the most biologically significant features (such as genes, isoforms, peptides and so on). 
Here, we are using a highly-configurable function "EnhancedVolcano" that produces publication-ready volcano plots.

```{r,eval=F,echo=F}
DEFindMarkers_Volcano <- function(DEFindMarkers) {
    # image: png


    # Changelog
    # 2022-09-14 Rearranged structure and description
    # 2020-10-29 Add support for pval == 0



    ## --------- ##
    ## Libraries ##
    ## --------- ##

    library(stringr)
    library(ggplot2)
    library(ggrepel)
    library(dplyr)
    library(tidyr)

    library(EnhancedVolcano)
    # For interactive plot:
    library(plotly)
    library(grid)
    

    ## -------------------------------- ##
    ## User-Defined Template Parameters ##
    ## -------------------------------- ##

    #Basic Parameters:
    df.orig <- DEFindMarkers
    label.col <- "Gene"
    sig.col <- c("C_HER2_vs_ER_pval","C_HER2_vs_TNBC_pval","C_ER_vs_TNBC_pval")
    pCutoff  = 0.001
    lfc.col <- c("C_HER2_vs_ER_logFC","C_HER2_vs_TNBC_logFC","C_ER_vs_TNBC_logFC")
    FCcutoff = 1.0
   
    
    #Label Parameters
    value_to_sort_the_output_dataset <- "p-value"
    no_genes_to_label <- 30
    use_only_addition_labels <- FALSE
    additional_labels <- ""
    is_red <- TRUE
    labSize <- 4    


    #Title and Axis labels Parameters
    change_sig_name <- "p-value"
    change_lfc_name <- "log2FC"
    title <- "p-value <= 0.001 & logFC >= 1.0"
    #subtitle <- ""
    use_custom_lab <- FALSE
            
    #Plot Parameters
    ylim <- 0
    custom_xlim <- ""
    xlim_additional <- 0
    ylim_additional <- 0
    axisLabSize <- 24
    pointSize <- 2


    #Image Parameters
    imageWidth = 3000
    imageHeight = 3000
    dpi = 300


  
    ##--------------- ##
    ## Error Messages ##
    ## -------------- ##

    ## --------------- ##
    ## Main Code Block ##
    ## --------------- ##

    rank <- list()
    for(i in 1:length(lfc.col)){
        lfccol <- lfc.col[i]
        sigcol <- sig.col[i]
        columns_of_interest <- c(label.col,lfc.col[i],sig.col[i])
        df <- df.orig %>% dplyr::select(one_of(columns_of_interest)) %>% 
                        mutate(!!sym(lfccol) := replace_na(!!sym(lfccol), 0)) %>%
                        mutate(!!sym(sigcol) := replace_na(!!sym(sigcol), 1)) 
                        #mutate(.data[[lfc.col[i]]] = replace_na(.data[[lfc.col[i]]], 0)) %>%
                        #mutate(.data[[sig.col[i]]] = replace_na(.data[[sig.col[i]]], 1)) 
        if (use_custom_lab==TRUE){
          if (nchar(change_lfc_name)==0){lfc_name = lfc.col[i]}
          if (nchar(change_sig_name)==0){sig_name = sig.col[i]}
          colnames(df) <- c(label.col,change_lfc_name, sig_name)
        } else {
          lfc_name = lfc.col[i]
          sig_name = sig.col[i]
        }
    
        group <- gsub("_pval|p_val_","",sig_name)
        rank[[i]] <- -log10(df[[sig_name]]) * sign(df[[lfc_name]]) 
        names(rank)[i] <- paste0("C_",group,"_rank")
        
        cat(paste0("Genes in initial dataset: ", nrow(df),"\n"))

        #Select top genes by logFC or Significance
            
        if (value_to_sort_the_output_dataset=="fold-change") {
            df <- df %>% dplyr::arrange(desc(.data[[lfc_name]]))
        } else if (value_to_sort_the_output_dataset=="p-value") {
            df <- df %>% dplyr::arrange(.data[[sig_name]]) 
        }

        if (is_red) {
            df_sub <- df[df[[sigcol]] <= pCutoff & abs(df[[lfccol]]) >= FCcutoff, ]
        } else {
            df_sub <- df
        }

        genes_to_label <- as.character(df_sub[1:no_genes_to_label,label.col])
#        additional_labels <- unlist(str_split(additional_labels,","))
 ## Modifying Additional Labels List:
 ## Replace commas with spaces and split the string
split_values <- unlist(strsplit(gsub(",", " ", additional_labels), " "))
additional_labels <- split_values[split_values != ""]

        filter <- additional_labels %in% df[,label.col]
        missing_labels <- additional_labels[!filter]
        additional_labels <- additional_labels[filter]

        if(length(missing_labels) > 0){
            cat("Could not find:\n")
            print(missing_labels)
        }

        if(use_only_addition_labels){
            genes_to_label <- additional_labels
        }else{
            genes_to_label <- unique(append(genes_to_label, additional_labels))
        }

        significant = vector(length = nrow(df))
        significant[] = "Not significant"
        significant[which(abs(df[,2]) > FCcutoff)] = "Fold change only"
        significant[which(df[,3] < pCutoff)] = "Significant only"
        significant[which(abs(df[,2]) > FCcutoff & df[,3] < pCutoff)] = "Significant and fold change"
        print(table(significant))
        
        # fix pvalue == 0
        shapeCustom <- rep(19,nrow(df))
        maxy <-  max(-log10(df[[sig_name]]), na.rm=TRUE)
        if(ylim > 0){
            maxy <- ylim
        }
      
        cat(paste0("Maxy: ",maxy,"\n"))
        if(maxy == Inf){
            # Sometimes, pvalues == 0
            keep <- df[[sig_name]] > 0
            df[[sig_name]][!keep] <- min(df[[sig_name]][keep])
            shapeCustom[!keep] <- 17

            maxy <- -log10(min(df[[sig_name]][keep]))
            cat("Some p-values equal zero. Adjusting y-limits.\n")
            cat(paste0("Maxy adjusted: ",maxy,"\n"))

        }

        # By default, nothing will be greater than maxy. User can set this value lower
        keep <- -log10(df[[sig_name]]) <= maxy
        df[[sig_name]][!keep] <- maxy
        shapeCustom[!keep] <- 17

        names(shapeCustom)<- rep("Exact",length(shapeCustom))
        names(shapeCustom)[shapeCustom == 17] <- "Adjusted"
        
        #Remove if nothin' doin'
        if(all(shapeCustom == 19)){
            shapeCustom <- NULL
        }
        
        maxy <- ceiling(maxy)

        if (grepl("log",lfc.col[i]) ){
                xlab <- bquote(~Log[2]~ "fold change")
        } else {
            xlab <- "Fold change"
        }
        if (grepl("adj",sig.col[i])){
            ylab <- bquote(~-Log[10]~ "FDR")
        } else {
            ylab <- bquote (~-Log[10]~ "p-value")
        }
        if(use_custom_lab){
            if(lfc_name != lfc.col[i]){
                xlab <- gsub("_"," ",lfc_name)
            }
            if (sig_name != sig.col[i]){ 
                ylab <- gsub("_"," ",sig_name)
            }
        }
  
## X-axis custom range change:
if (custom_xlim == "") {
        xlim=c(floor(min(df[,lfc_name])) - xlim_additional,ceiling(max(df[,lfc_name]))+ xlim_additional)
} else if (grepl(",", custom_xlim) == FALSE) {
    xlim=c(-1*as.numeric(trimws(custom_xlim)), as.numeric(trimws(custom_xlim)))
} else {
    split_values <- strsplit(custom_xlim, ",")[[1]]

    # Trim whitespace and convert to numeric values
    x_min <- as.numeric(trimws(split_values[1]))
    x_max <- as.numeric(trimws(split_values[2]))

    xlim <- c(x_min, x_max)
}


        p <- EnhancedVolcano( df,x=lfc_name,y=sig_name,
                              lab=df[,label.col],
                              selectLab = genes_to_label,
                              title=title, #CHANGE NW: See line 78
                              subtitle <- group,
                              xlab=xlab,
                              ylab=ylab,
                              xlim=xlim,
                              ylim=c(0, maxy + ylim_additional),
                              pCutoff=pCutoff,
                              FCcutoff=FCcutoff,
                              axisLabSize=axisLabSize,
                              labSize=labSize,
                              pointSize=pointSize,
                              shapeCustom=shapeCustom
                              )
        print(p)   
 ## Adding interactive plot with no labels:
        p_empty <- EnhancedVolcano( df,x=lfc_name,y=sig_name,
                              lab = rep("", nrow(df)),  # Setting labels to empty strings
                              selectLab = NULL,
                              title=title, #CHANGE NW: See line 78
                              subtitle <- group,
                              xlab=xlab,
                              ylab=ylab,
                              xlim=xlim,
                              ylim=c(0, maxy + ylim_additional),
                              pCutoff=pCutoff,
                              FCcutoff=FCcutoff,
                              axisLabSize=axisLabSize,
                              labSize=labSize,
                              pointSize=pointSize,
                              shapeCustom=shapeCustom
                              )
##        print(p_empty)   

# Extract the data used for plotting
plot_data <- ggplot_build(p_empty)$data[[1]]

pxx <- p_empty +
  xlab("Fold Change") +  # Simplify x-axis label
  ylab("Significance") + # Simplify y-axis label
  theme_minimal() +
  geom_point(aes(
    text = paste("Gene:", df[[label.col]], 
                 "<br>Log2FC:", df[[lfc_name]], 
                 "<br>P-value:", df[[sig_name]]),
    colour = as.character(plot_data$colour),
    fill = as.character(plot_data$colour)  # Set fill to the same as colour
  ), 
  shape = 21,  # Shape that supports both colour and fill
  size = 2,  # Size of the points
  stroke = 0.1  # Stroke width
  ) + scale_fill_identity()

# Add interactive hover labels for the gene names
interactive_plot <- ggplotly(pxx, tooltip = c("text"))


    ## showing interactive plot
    grid.newpage()
    print(interactive_plot)
    grid.newpage()
    ## The end of addition
   
    }
   
    df.final <- cbind(df.orig, do.call(cbind, rank))
    return(df.final)
}
```


1. Maria Doyle, 2021 Visualization of RNA-Seq results with Volcano Plot (Galaxy Training Materials). https://training.galaxyproject.org/training-material/topics/transcriptomics/tutorials/rna-seq-viz-with-volcanoplot/tutorial.html Online; accessed Mon Aug 01 2022
2. Batut et al., 2018 Community-Driven Data Analysis Training for Biology Cell Systems 10.1016/j.cels.2018.05.012
3. Blighe, K, S Rana, and M Lewis. 2018. EnhancedVolcano: Publication-ready volcano plots with enhanced coloring and labeling. https://github.com/kevinblighe/EnhancedVolcano.





