---
title: "Differential Expression Analysis"
author: "CCBR"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SCWorkflow-DEG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE, message = FALSE
)

library(dplyr)


run_Chunks=TRUE
```

```{r include=F,echo=F,eval=run_Chunks}
library(Seurat)
library(data.table)
library(ggplot2)
library(plotly)

# Source required functions explicitly for interactive vignette execution
deg_functions <- c("DEG_Gene_Expression_Markers.R", "AggregateCounts.R", "Volcano_Plot.R")
invisible(sapply(deg_functions, function(f) {
  tryCatch(source(file.path("../R", f)), error = function(e) {
    warning("Could not source ", f, ": ", e$message)
  })
}))

```

```{r,include=F,echo=F,eval=run_Chunks}

Anno_SO_raw=readRDS("./images/Anno_SO.Seurat5.rds")
Anno_SO=if (inherits(Anno_SO_raw, "Seurat")) Anno_SO_raw else Anno_SO_raw$object

```

## DE with Find Markers 

---

This function performs a DE (differential expression) analysis on a merged Seurat object to identify expression markers between different groups of cells (contrasts). This analysis uses the FindMarkers() function of the Seurat Workflow. 

**Methodology**
Differential expression analysis (DEG) is a fundamental technique in single-cell genomics research. The goal of DEG analysis is to identify genes that exhibit significant changes in expression levels between different groups of cells or conditions, thereby uncovering potential markers that distinguish these groups.
This function takes a merged Seurat object [1] as input, expected to contain single-cell data from multiple samples, along with relevant metadata and SingleR annotations, which provide information about cell identity. To perform the DEG analysis, the user can choose from various statistical algorithms, such as MAST [2], wilcox [3], bimod [4], and more, which accommodate different types of experimental designs and assumptions about the data. The user can control the sensitivity of the analysis by setting the minimum fold-change in gene expression between the groups to be considered significant. Additionally, users can specify the assay to be used for the analysis, whether it is the scaled data (SCT) or raw RNA counts.
For best results, it is recommended to use this function with well-curated and preprocessed single-cell data, ensuring that the Seurat object contains relevant metadata and SingleR annotations. Users should carefully select the samples and contrasts based on their experimental design and research questions. Additionally, exploring different statistical algorithms and adjusting the threshold can fine-tune the DEG analysis and reveal more accurate gene expression markers.


```{r,eval=run_Chunks}

DEG_table=degGeneExpressionMarkers(object = Anno_SO,
                                   samples = c("PBS", "ENT", "NHSIL12", "Combo", "CD8dep" ),
                                   contrasts = c("0-1"),
                                   parameter.to.test = "SCT_snn_res_0_2",
                                   test.to.use = "MAST",
                                   log.fc.threshold = 0.25,
                                   assay.to.use = "SCT",
                                   use.spark = F
                                   )

```

```{r echo=F}
if (run_Chunks==T) {
  deg_table_df <- DEG_table$data$DEG_Table
  write.table(deg_table_df,
            file = "./images/DEG_Table.knit-test.txt",
            sep = '\t',
            row.names = F)
}else{
  deg_table_df <- read.delim("./images/DEG_Table.knit-test.txt")
}

  deg_table_df %>% head() %>% knitr::kable()

```

1. https://satijalab.org/seurat/
2. https://rglab.github.io/MAST/
3. Dalgaard, Peter (2008). Introductory Statistics with R. Springer Science & Business Media. pp. 99–100
4. https://en.wikipedia.org/wiki/Multimodal_distribution

## Pseudo Bulk Method

---

### Aggregate Seurat Counts

This function is the first step in a Pseudobulk analysis of your scRNA-seq dataset. It groups cells based on chosen categorical variable(s) in the Seurat Object's Metadata and aggregates the counts of each gene in each group.

The output is a table of aggregate expression in which the rows are genes and the columns are values found in the chosen Pseudobulk variable. If you select multiple categories to aggregate by (e.g. Category1: A,B,C and Category2: D,E,F), cells will be grouped by combinations of category variables (e.g. A_D, A_E, A_F, B_D, B_E, B_F). By default, gene counts are averaged across cells in each group.

```{r,eval=TRUE,echo=FALSE}

aggregateCounts=function(object,
                         var.group,
                         layer){
  
  
  ## --------------- ##
  ## Main Code Block ##
  ## --------------- ##
  
  pseudobulk <- AverageExpression(object, 
                                  return.seurat = FALSE,
                                  assay = "SCT",
                                  group.by = var.group,
                                  layer = layer)[[1]] %>% 
                  as.data.frame.matrix() 
    
  pseudobulk$Gene <- rownames(pseudobulk)
  pseudobulk <- pseudobulk %>% select("Gene", everything())
  rownames(pseudobulk) <- NULL

  # Further processing of column names
  colnames(pseudobulk) <- gsub("\\W","_",colnames(pseudobulk))

  # # Rename columns generated from Clustering to be more Generic
  #   if(grepl("SCT_snn_res",var.group)){ 
  #     colnames(pseudobulk)[grepl("\\d+", colnames(pseudobulk))] <- 
  #       paste("Cluster", 
  #             colnames(pseudobulk)[grepl("\\d+", colnames(pseudobulk))], 
  #             sep = "_")
  #   }
  
  # Return Table/Figure that gives statistics on 
  # Number of Cells in each group/new sample
  # Distribution of cell Counts in each group/new sample
  meta <- object@meta.data[,var.group]
  
  # check that columns are all factors / categorical
  char_or_factor_cols <- sapply(meta, function(x) is.character(x) || is.factor(x) || is.logical(x))
  
  # do plots and tables if all columns are factors or characters
  if(all(char_or_factor_cols)){
    meta$interaction <- gsub("\\W","_",interaction(meta))
    
    df <- as.data.frame(table(pseudobulk_group = meta$interaction)) %>% filter(Freq != 0)
    # sort the table by the number of cells in each group
    df <- df[order(df$Freq, decreasing = F),]
    
    if(any(df$Freq == 1)){
      single_counts <- df$pseudobulk_group[df$Freq == 1]
      # sprintf, make custom warning message with %s as placeholder for single count groups
      warning(sprintf(
        "Some groups have only 1 cell. It is recommended to have at least 2 cells in each group.\nAffected groups: %s",
        paste(single_counts, collapse = ", ")
      ))
    }
    
    p <- ggplotly(ggplot(df, aes(x = pseudobulk_group, y = Freq)) +
                    geom_bar(stat = "identity", position = "stack") +
                    labs(y = "Counts", x = "Pseudobulk Groups", title = "Number of Cells in each Pseudobulk Group") +
                    theme(axis.text.x = element_text(angle = 90, hjust = 1)))
    
    print(p)
    
  } else {
    stop("All columns in var.group must be factors or characters")
  }
  
  return(pseudobulk)
}
```

```{r,eval=F}
aggregateCounts(object=so,
               var.group=var_group,
               layer=layer)
```


### Statistical Analysis using Limma

Given a matrix (typically log-normalized gene expression) and a metadata table, this will run one- and two-factor statistical analyses on groups using linear or mixed effects models with limma. Reference.

There are 2 ways for treating Donor or Patient - one as a random effect and the other as a fixed effect

When Using a Mixed Effects Model (Donor as random effect):

Add Donor column to the Donor Variable Column
Do not add the Donor variable under Covariate Columns. It will be handled separately in the Donor Variable Column as a random effect.
The Covariate Columns field should include any other variables except the Donor.
When Using a Basic Linear Model (Donor as fixed effect):

You can add the Donor column as a covariate under Covariate Columns, where it will be treated as a fixed effect.
Additional variables can be included under Covariate Columns
Ensure the Donor Variable Column is left blank.
This function is a Beta version and is undergoing active development. If you encounter problems, please contact CCBR NCICCBRNIDAP@mail.nih.gov


```{r,eval=F}
Pseudobulk_LimmaStats()
```

## Visualizations

---

### Volcano Plot - Enhanced

This function utilizes the *EnhancedVolcano* R Bioconductor package to generate publication-ready volcano plots for differential expression analyses, offering a number of customizable visualization options and optimizing gene label placement to avoid clutter


**Methodology**
A volcano plot is a type of scatterplot that shows statistical significance (P value) versus magnitude of change (fold change). It enables quick visual identification of genes with large fold changes that are also statistically significant. These may be the most biologically significant features (such as genes, isoforms, peptides and so on). 
Here, we are using a highly-configurable function "EnhancedVolcano" that produces publication-ready volcano plots.

```{r, eval=run_Chunks, echo=FALSE}
deg_table_df <- if (exists("deg_table_df")) {
  deg_table_df
} else if (is.list(DEG_table) && "data" %in% names(DEG_table) && "DEG_Table" %in% names(DEG_table$data)) {
  DEG_table$data$DEG_Table
} else {
  DEG_table
}

lfc_cols <- grep("_logFC$", colnames(deg_table_df), value = TRUE)
sig_cols <- grep("_pval$|^p_val_", colnames(deg_table_df), value = TRUE)

normalize_name <- function(x) {
  x <- gsub("_logFC$", "", x)
  x <- gsub("_pval$", "", x)
  x <- gsub("^p_val_", "", x)
  x
}

lfc_keys <- normalize_name(lfc_cols)
sig_keys <- normalize_name(sig_cols)
matched <- intersect(lfc_keys, sig_keys)

if (length(matched) == 0) {
  stop("No matching logFC/p-value column pairs were found in DEG output")
}

lfc_use <- lfc_cols[match(matched, lfc_keys)]
sig_use <- sig_cols[match(matched, sig_keys)]

volcano_out <- Volcano_Plot(
  DEGAnalysis = deg_table_df,
  label.col = "Gene",
  sig.col = sig_use,
  lfc.col = lfc_use,
  pCutoff = 0.001,
  FCcutoff = 1.0,
  title = "p-value <= 0.001 & logFC >= 1.0",
  axisLabSize = 11,
  titleLabSize = 10,
  subtitleLabSize = 8,
  captionLabSize = 6,
  legendLabSize = 8,
  labSize = 3,
  pointSize = 1.0
)

plot_keys_static <- names(volcano_out)[grepl("_static$", names(volcano_out))]
plot_keys_interactive <- names(volcano_out)[grepl("_interactive$", names(volcano_out))]

if (length(plot_keys_static) > 0) {
  for (k in plot_keys_static) {
    print(volcano_out[[k]])
  }
}

if (length(plot_keys_interactive) > 0) {
  for (k in plot_keys_interactive) {
    print(volcano_out[[k]])
  }
}
```


1. Maria Doyle, 2021 Visualization of RNA-Seq results with Volcano Plot (Galaxy Training Materials). https://training.galaxyproject.org/training-material/topics/transcriptomics/tutorials/rna-seq-viz-with-volcanoplot/tutorial.html Online; accessed Mon Aug 01 2022
2. Batut et al., 2018 Community-Driven Data Analysis Training for Biology Cell Systems 10.1016/j.cels.2018.05.012
3. Blighe, K, S Rana, and M Lewis. 2018. EnhancedVolcano: Publication-ready volcano plots with enhanced coloring and labeling. https://github.com/kevinblighe/EnhancedVolcano.





